Writing Trash Bin

### 2024-02-23 paper.qmd
finished the task wordvomit and future work  sections

---
title: "Noisy Work: A Review of The Uncertainty Visualisation Literature"
author: Harriet Mason
bibliography: references.bib
date: last-modified
format: pdf
editor_options: 
  chunk_output_type: console
---

# 1. Introduction
::: aside
I have not touched this section since the confirmation so it would be best to ignore it
:::

@@

## 3.1 Coloquial Definitions
::: aside
I am not sure how to collect these together and show that they are highly varied. This might be a situation for the tidytext clustering or something. There is an additional problem is that the definition for uncertainty is *completely* unrelated to the uncertainty that is quantified in the graphics. I think the tidytex can help with the definitions being all over the place, but I am not sure how to identify the lack of correlation between the colloquial definitions of uncertainty and the mathematical quantificaitons.
:::

@@

## 3.2 Taxonomy Definitions
### 3.2.1 Taxonomies of Uncertainty
::: aside
I actually think I agree with the sentiment in the grammar of graphics that a taxonomy is more of an opinion and we can reclassify it and is therefore not thorough enough to remain unchanged and be build upon. That being said, if this is a literature review and all existing literature use written definitions or taxnomies, that is what I have to work with. The larger taxonomy can be used to show that the other taxonomies fit within it, but it is in of itself not tied to anything except work *that paper itself* identifies as transient.
:::

@@

::: aside
This figure is not going to be in the final version of the paper. It is here as a placeholder. Ideally, I would replace with with a graphic that shows where all these taxonomies land in the location, level, nature taxonomy (since it is encompassing) and show they are very inconsistent and a new definition is pulled out for almost every paper (even in statistics papers)
:::

@@

::: aside
I had this here to explain why these are the considerations people make in statistics and how they relate to elements of the taxonomy to try and provide some justification, but If I am just going to show the definitions are inconsitent, I might get rid of this section. I also largely included it so you guys can see what the taxonomies are considering
:::

@@

::: aside
I should add something in about the actual quantified uncertainty vs the discussed uncertanty thing (i.e. what is defined with words and what is defined with maths)
:::

@@

*this is all currently from the spieglehatter paper because that is the only paper I have moved notes over from mendeley before I started doing the summary write up. this will not all be from one paper, dont worry.*

::: aside
This section still needs some citations from papers that communicate uncertainty, but the are *hard* to organise because of the ambiguous uncertainty definition problem. So they will all be communicating something different but you cant find out what it is until you are knee deep in the method section.
:::

@@

::: aside
I should maybe change this to relatability to uncertainty, value extraction, decision making and trust, and meta papers or something. Maybe perception and evaluation papers?
:::

@@

## Uncertainty and Task Specificity
When we consider mapping uncertainty, a reasonable questions that should come to mind is "uncertainty about what?". Every single thing concept or activity in statistics does not have uncertainty. For example, descriptive statistics cannot have uncertainty as they are just describing statistics as they are, the sample mean is the sample mean, we are not uncertain about its value. A sample variance describes dispersion, which is also how we describe uncertainty, but the sample variance in of itself does not have any uncertainty. Most of the mathematical uncertainty concepts have this idea embedded in them. For example, a confidence interval is an interval drawn around the sample mean that will theoretically contain the *populaton mean* 95% of the time. This issue has been noticed several times, however it is often brought up as a *task* dependence rather than an understanding that uncertainty is inherrently linked to inference on a particular statistic. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failling to do it well can cause downstream effects and damage the effectivness of a visualisation. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. This should make it clear that uncertainty only exists when we are performing inference. Therefore it only makes sense to describe uncertainty in terms of something to be uncertain *about*, it is not some a latent feature of statistics and data.

@@

- @Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included". @Potter2010 aimed to create a summary plot that "concisely presented data with uncertaninty information". These papers fail to realise the exploration step of an analysis, which includes descriptive statistics, exploratory data visualisaiton, and unsupervised machine learning techniques, is performed without a prior hypothesis. Therefore, it is performed to *suggest* a direction for future inference work or communication, which *do* contain uncertainty and does not contain uncertainty within itself. This confusion would not occur if it was understood that uncertainty does not exist without inference.

@@

This misunderstanding is not due to laziness or the fault of the authors, but rather is likely caused by the absence of a strict definition of uncertainty. The imprecise and confusing set of existing definitions are creating a field in which the authors themselves do not know what they are testing. They do not even seem to understand what uncertainty is. Uncertainty examples include imputed data, model selection, inherrent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference. These things are not "uncertainty" in of themselves, but rather contribute to the distance between the final estimate and the statistic we want to perform inference on. The field of uncertainty communication is in desperate need of a unifying mathematical definition that will allow the bias and variance introduced through various stages of the data analysis pipeline to be compared or quantified.

Most uncertainty definitions speak about this vague concept of uncertainty but do not

The fact that uncertainty only exists when we are performing inference also creates issues with the questions of an experiment. If uncertainty is depicted and uncertainty is always related to . The authors of these experiments often comment on the results from the perspective that the participants have made a conceptual mistake, and at no point do they question why we would depict uncertainty for an unrelated ground truth. The results of these experiments essentially show that participants can be "tricked" by failling to understand uncertainty (showing that people can be tricked if you purposely lead them astray should not constitute interesting research). Ironically the authors fail to realise they have made this same mistake when they did not consider displaying uncertainty for a fundamentally unrelated

assume the participants have made some sort of mistake, rather than questioning their own understanding of uncertainty.

must be related to the statistic we wish to draw inference on

Reguardless of whether or not the visualisation and associated uncertainty is related to the statistic we are trying to draw inference from, participants will *assume* they are.

@Hofman2020 suggests authors depictions of the

This concept is the first issue that uncertainty experiments run into, specifically in perceptual task experiments.

It is rarely mentioned in the uncertainty literature, but

The choice of what we map uncertainty to will impact how easily

Even the most simple experiments are not free from the issues in the uncertainty definition.
Some experiments directly relate the internal

-  Error is best mapped to fuzziness, location, and colour value; arrangement, size and transparency are an OK second choice; but saturation, hue, orientation and shape are unacceptable and have no intuitive connection to variance [@Maceachren2012]. No only do the graphical elements we map our features to matter, but the direction matters too. Graphical elements that are more fuzzy (fuzziness), further from centre (location), lighter (colour value), poorly arranged (arrangement), smaller (size), more transparent (transparency) are perceived to be more uncertain [@Maceachren2012]. This idea also extends to interval estimation, where questions about probability are best answered with gradients and questions about start and end times are easiest to answer with ambiguation [@Gschwandtnei2016]. Heuristics can work against us just as much as they can work for us. The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015]

- @Nathonours found that a scatter plot is better than a line plot if you want to convey the correlation between two time series but we cannot be sure if this was influenced by swapping the distribution depicted or by dropping the irrelevant feature of inexchangeability.

- I think the dimension of the uncertainty should be considered every time there is aggregation from multiple sources of uncertainty and you want to maintain those sources. You can project the uncertainty down into one dimension if the source information is in of itself unimportant.

@@

::: aside
I think I want the "in order to visualise uncertainty you need to quantify it and the quantification does not align with the definition" point here since this section is now about value extraction.
:::

@@

A bizarre feature of uncertainty visualisation literature is that it does not work to build upon these existing principle or identify the ways in which uncertainty visualisations may diverge from these rules, but rather it seems to exist with complete indifference to the research in information visualisation at large. The most common uncertainty visualisation evaluation study asks participants to perform some value abstraction either explicitly by asking participants to report values or vaguely by asking participants to find regions of least certainty, rank plots, report their confidence, and/or make decisions according to uncertainty information. Experimenters then compare these results based on the participants accuracy, preference, confidence, and/or speed @Hullman2016. Most commonly, these value abstraction papers compare two different visualisations of uncertainty on a set of these questions, however, since there is no rules for what is or is not an uncertainty visualisation the graphics that are compared often differ in both computation *and* information. Uncertainty visualisaiton papers only seem to require that the plots depict "uncertainty" which has already been established to be poorly defined. The complete lack of formality is easy to see if we actually plot the visualisations compared across evaluation experiments according to the grammar of graphics *in the figure below*:

**INCLUDE A GRAPHIC THAT SHOWS WHERE EACH TYPE OF UNCERTAINTY VISUALISATION SITS IN THE GRAMMAR OF GRAPHICS AND HOW DIFFERENT THEY ARE ACROSS MULTIPLE STUDIES**

@@

@@

# 5 Great Examples
- best things you have seen and why they are the best
- can't have a best practices because you need experimental results and my point is that the experiments aren't testing the right things
- pixel map
  - Similar to what could be good but its computation isnt quite what I would like
  - heaps of sample depictions are animated and that is not always possible, but this is a clear indicator of how you can do a sample visualisaiton
  - hasnt been tested so it is unclear if this actually works at signal supression
- VSUP
  - The purpose of uncertainty is clear in the design and it is signal supression
  - Supression is not the only uncertainty feature we should be interested in (shape is also important which is why I have beef with over smoothing and something visualisation is very good to express) but by being clear in its goal it knows what it is
  - It addresses several key issues with the bivariate map that arrise in general and especially when it is used for visualising uncertainty 
- the census dot map (has been taken down but CNN replicated it [here](https://edition.cnn.com/interactive/2021/us/census-race-ethnicity-map/))
  - I love a good depiction of a sample
  - You can see the sparse regions and the pop 
  - it is similar to the pixel map (could also be singal supression but it is not tested)
  - the sample depiction works well with the interactivity and you can see the signal that is appropriate for a particular level of zoomed in. If you zoom out the points turn into solid colours and show clear signal.
- The new york times class mobility and race animated plot [here](https://www.nytimes.com/interactive/2018/03/19/upshot/race-class-white-and-black-men.html) 
- the animated surface classifications from that satelite images experimental paper
- the climate change scenario uncertainty. There are quite a few of these:
  - [e.g.1](https://earthobservatory.nasa.gov/features/GlobalWarming/page5.php)
  - [e.g.2](https://climatechange.chicago.gov/sites/production/files/2016-07/scenariotempgraph_0.jpg)
  - there are a few examples on the [climate change scenario wiki](https://en.wikipedia.org/wiki/Climate_change_scenario)
  - Good example of uncertainty that has uncertainty from two different sources (human decision in the form of inputs and statistical uncertainty), one of which cant be quantified.
  - a good example in how you can combine two different  types of uncertainty
  - [This](https://www.esri.com/arcgis-blog/products/arcgis-pro/mapping/graphical-cartograms-in-arcgis-pro/) has some good maps and some bonkers ones 
- Exceedance probability map (example where uncertainty that *should* be conveyed as signal is expressed as a signal. It is a typical maths problem that involves)
Notes
- Since a lot of the good visualisations are made by random people creating a new visualisation (likely through back and forth communication which is HOW you get good visualisations) 
- A lot of these show the data instead of an inference which is probably better for this scenario because sample size is important for uncertainty. It is not always appropriate and sometimes dots are not exactly equal to one draw. 
  - comments about smoothing in density plots and bin width in histograms. I do think the sample size should be communicated to people. While it is not the only thing that matters it seems kind of pointless to go to such lengths to hide it.
- Sometimes points are used for things that are NOT samples which can get confusing and honestly I dont like it.

# 5. Future work/Conclusion

# Appendix (or information I am holding)
## Experiment paper checklist
General
- Specific Field of Research
- Question asked
- Risk or Uncertainty: 
- Noise vs signal (risk vs uncertainty)
- Major task goal
    - e.g. identify a bias or misunderstanding, compare two visualisation methods
- Minor task goal
    - Communication(value retrieval)
    - Analytical(complex consideration)
    - Exploration(free for all)

Uncertainty Considerations
	- Source
	- Level (statistical-ignorance)
	- Nature (Epistemic/Aleatory)

Visual Uncertainty taxonomy
- Explicit/implicit
    - (directly mapping or showing multiple outcomes)
- Intrinsic/Extrinsic 
    - (using existing symbols e.g. colour value, or new objects e.g. grids)
- Visually integral/separable 
    - (can or cannot be separated from the data) 
- Coincidence/adjacent 
    - (if data and uncertainty are in integrated or separate views
- Static/dynamic 
    - (animation/interaction)

Other plot considerations
	- Dimension of data
	- Dimension of uncertainty
	- Feature mappings

Extra (not necessarily recorded)
	- Metrics used (and recorded)
	- Possible Heuristics
	- What is the ground truth?
	- Participant literacy

>>>>

---
title: "Noisy Work: A Review of The Uncertainty Visualisation Literature"
author: Harriet Mason
bibliography: references.bib
date: last-modified
toc: true
number-sections: true
format: pdf
editor_options: 
  chunk_output_type: console
---

# Introduction

**Aside: I have not touched this section since the confirmation so it would be best to ignore it**

@@

## 3.1 Coloquial Definitions
**Aside: I am not sure how to collect these together and show that they are highly varied. This might be a situation for the tidytext clustering or something. There is an additional problem is that the definition for uncertainty is completely unrelated to the uncertainty that is quantified in the graphics. I think the tidytex can help with the definitions being all over the place, but I am not sure how to identify the lack of correlation between the colloquial definitions of uncertainty and the mathematical quantificaitons.**

@@

## 3.2 Taxonomy Definitions
### 3.2.1 Taxonomies of Uncertainty
**Aside: I actually think I agree with the sentiment in the grammar of graphics that a taxonomy is more of an opinion and we can reclassify it and is therefore not thorough enough to remain unchanged and be build upon. That being said, if this is a literature review and all existing literature use written definitions or taxnomies, that is what I have to work with. The larger taxonomy can be used to show that the other taxonomies fit within it, but it is in of itself not tied to anything except work that paper itself identifies as transient.**

@@

**Aside: This figure is not going to be in the final version of the paper. It is here as a placeholder. Ideally, I would replace with with a graphic that shows where all these taxonomies land in the location, level, nature taxonomy (since it is encompassing) and show they are very inconsistent and a new definition is pulled out for almost every paper (even in statistics papers)**

@@

**Aside: I had this here to explain why these are the considerations people make in statistics and how they relate to elements of the taxonomy to try and provide some justification, but If I am just going to show the definitions are inconsitent, I might get rid of this section. I also largely included it so you guys can see what the taxonomies are considering**

@@

**Aside: I should add something in about the actual quantified uncertainty vs the discussed uncertanty thing (i.e. what is defined with words and what is defined with maths)**

@@

**Aside:this is all currently from the spieglehatter paper because that is the only paper I have moved notes over from mendeley before I started doing the summary write up. this will not all be from one paper, dont worry.**

**Aside:This section still needs some citations from papers that communicate uncertainty, but the are *hard* to organise because of the ambiguous uncertainty definition problem. So they will all be communicating something different but you cant find out what it is until you are knee deep in the method section.**

@@

**Aside: I should maybe change this to relatability to uncertainty, value extraction, decision making and trust, and meta papers or something. Maybe perception and evaluation papers?**

@@

## Uncertainty and Inference
When we consider mapping uncertainty, a reasonable questions that should come to mind is "uncertainty about what?". Every single thing concept or activity in statistics does not have uncertainty. For example, descriptive statistics cannot have uncertainty as they are just describing statistics as they are, the sample mean is the sample mean, we are not uncertain about its value. A sample variance describes dispersion, which is also how we describe uncertainty, but the sample variance in of itself does not have any uncertainty. Most of the mathematical uncertainty concepts have this idea embedded in them. For example, a confidence interval is an interval drawn around the sample mean that will theoretically contain the *populaton mean* 95% of the time. This issue has been noticed several times, however it is often brought up as a *task* dependence rather than an understanding that uncertainty is inherrently linked to inference on a particular statistic. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failling to do it well can cause downstream effects and damage the effectivness of a visualisation. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. This should make it clear that uncertainty only exists when we are performing inference. Therefore it only makes sense to describe uncertainty in terms of something to be uncertain *about*, it is not some a latent feature of statistics and data.

@@

- @`Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included". @Potter2010 aimed to create a summary plot that "concisely presented data with uncertaninty information". These papers fail to realise the exploration step of an analysis, which includes descriptive statistics, exploratory data visualisaiton, and unsupervised machine learning techniques, is performed without a prior hypothesis. Therefore, it is performed to *suggest* a direction for future inference work or communication, which *do* contain uncertainty and does not contain uncertainty within itself. This confusion would not occur if it was understood that uncertainty does not exist without inference.

@@

This misunderstanding is not due to laziness or the fault of the authors, but rather is likely caused by the absence of a strict definition of uncertainty. The imprecise and confusing set of existing definitions are creating a field in which the authors themselves do not know what they are testing. Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty *when we perform inference*. These things are not "uncertainty" in of themselves, but rather contribute to the distance between the final estimate and the statistic we want to perform inference on. The field of uncertainty communication is in desperate need of a unifying mathematical definition that firmly identifies exactly how each of these things contribute to uncertainty. This may be a considerable task, but as of right now, the list of things that are "uncertainty" continues to grow, and the best methods of uncerstanding and quantifying them are ad-hoc at best. @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, @Meng2014 mathematically defined the variance introduced to a model by the array of model choices, information theory tries to quantify uncertainty using the idea of entropy. None of these existing methods are thorough enough for an analyst to understand what causes uncertainty, and quantify it for communication. This limitation becomes apparent when you look at the limited scope of what the uncertainty visualisation literatrue has *actually* focused on visualising. Almost every uncertainty visualisation paper contains a broad and encapsulating definition of uncertainty in the introduction, followed by a visualisation of a normally distributed probablity density function.

**Aside: Include table that shows broad uncertainty definition and then PDF visualisation**

In order to visualise something, it needs to be quantifiable. In order to quanitfy something, it needs to be mathematically defined. The broad classes of what is consdiered "uncertainty" is not currently quantifiable in any way that is not ad-hoc. The only apsects of uncertaitny that are currently quantifiable are confidence intervals, prediction intervals, and related terms. Even then, these quantifications often only capture the uncertainty in our model or its assumptions, it ignores bias often introduced in earlier stages of the analysis. Broader and more complicated concepts such as the effects of assumptions, imputed missing variables, and model choices remain difficult to meaningfully quantify beyond ad-hoc methods.  This means that the current literature of uncertainty visualisation papers are doing one of two things:  
1) Evaluating multiple visual representations of a single mathematical expression of uncertainty (e.g. a PDF, confidence interval or error). These papers are purposefully vague, and would be clearer in their intention if the title was aligned with the actual mathematical object they are visualising. For example, a paper that is titled "evaluation of uncertainty visualisations" that only considers PDF visualisation should realistically be titled "evaluation of PDF visualisations".
2) Evaluating multiple visual representations of *multiple* different mathematical expression of uncertainty. Since these experiments visualise different mathematical objects, some of which are closer to the ground truth statistic, the results of these papers can be easily predicted with an understanding of how to get the ground truth statistic from a plot and the perceptual tasks required to do so.

It is clear that the vague definition is causing across the board misunderstandings of uncertainty, confusing experiment methodologies and motivations, and a literature that is difficult to navigate. This is also causing a large body of experiments that have easily predictable results that can be easily explained by the existing visualisation. This problem will be the focus of the following section.

@@

**Aside: I think I want the "in order to visualise uncertainty you need to quantify it and the quantification does not align with the definition" point here since this section is now about value extraction.**

@@

A bizarre feature of uncertainty visualisation literature is that it does not work to build upon these existing principle or identify the ways in which uncertainty visualisations may diverge from these rules, but rather it seems to exist with complete indifference to the research in information visualisation at large. The most common uncertainty visualisation evaluation study asks participants to perform some value abstraction either explicitly by asking participants to report values or vaguely by asking participants to find regions of least certainty, rank plots, report their confidence, and/or make decisions according to uncertainty information. Experimenters then compare these results based on the participants accuracy, preference, confidence, and/or speed @Hullman2016. Most commonly, these value abstraction papers compare two different visualisations of uncertainty on a set of these questions, however, since there is no rules for what is or is not an uncertainty visualisation the graphics that are compared often differ in both computation *and* information. @Hullman2018 asks participants questions that would require a prediction interval but provided a sampling distribution and argued that the two could be considered equivalent so long as participants understood the difference between a confidence interval and a prediction interval, knew the sample size, and knew the formula to convert the confidence interval to a prediction interval. The authors themselves note that this is rediculous, which begs the question. Uncertainty visualisaiton papers only seem to require that the plots depict "uncertainty" which has already been established to be poorly defined. The complete lack of formality is easy to see if we actually plot the visualisations compared across evaluation experiments according to the grammar of graphics *in the figure below*:

**Aside: INCLUDE A GRAPHIC THAT SHOWS WHERE EACH TYPE OF UNCERTAINTY VISUALISATION SITS IN THE GRAMMAR OF GRAPHICS AND HOW DIFFERENT THEY ARE ACROSS MULTIPLE STUDIES**

@@

- @Patrick2023 investigated peoples ability to identify a signal and compared it to typical hypothesis tests

@@

# Great Examples
- best things you have seen and why they are the best
- can't have a best practices because you need experimental results and my point is that the experiments aren't testing the right things
- pixel map
  - Similar to what could be good but its computation isnt quite what I would like
  - heaps of sample depictions are animated and that is not always possible, but this is a clear indicator of how you can do a sample visualisaiton
  - hasnt been tested so it is unclear if this actually works at signal supression
- VSUP
  - The purpose of uncertainty is clear in the design and it is signal supression
  - Supression is not the only uncertainty feature we should be interested in (shape is also important which is why I have beef with over smoothing and something visualisation is very good to express) but by being clear in its goal it knows what it is
  - It addresses several key issues with the bivariate map that arrise in general and especially when it is used for visualising uncertainty 
- the census dot map (has been taken down but CNN replicated it [here](https://edition.cnn.com/interactive/2021/us/census-race-ethnicity-map/))
  - I love a good depiction of a sample
  - You can see the sparse regions and the pop 
  - it is similar to the pixel map (could also be singal supression but it is not tested)
  - the sample depiction works well with the interactivity and you can see the signal that is appropriate for a particular level of zoomed in. If you zoom out the points turn into solid colours and show clear signal.
- The new york times class mobility and race animated plot [here](https://www.nytimes.com/interactive/2018/03/19/upshot/race-class-white-and-black-men.html) 
- the animated surface classifications from that satelite images experimental paper
- the climate change scenario uncertainty. There are quite a few of these:
  - [e.g.1](https://earthobservatory.nasa.gov/features/GlobalWarming/page5.php)
  - [e.g.2](https://climatechange.chicago.gov/sites/production/files/2016-07/scenariotempgraph_0.jpg)
  - there are a few examples on the [climate change scenario wiki](https://en.wikipedia.org/wiki/Climate_change_scenario)
  - Good example of uncertainty that has uncertainty from two different sources (human decision in the form of inputs and statistical uncertainty), one of which cant be quantified.
  - a good example in how you can combine two different  types of uncertainty
  - [This](https://www.esri.com/arcgis-blog/products/arcgis-pro/mapping/graphical-cartograms-in-arcgis-pro/) has some good maps and some bonkers ones 
- Exceedance probability map (example where uncertainty that *should* be conveyed as signal is expressed as a signal. It is a typical maths problem that involves)
Notes
- Since a lot of the good visualisations are made by random people creating a new visualisation (likely through back and forth communication which is HOW you get good visualisations) 
- A lot of these show the data instead of an inference which is probably better for this scenario because sample size is important for uncertainty. It is not always appropriate and sometimes dots are not exactly equal to one draw. 
  - comments about smoothing in density plots and bin width in histograms. I do think the sample size should be communicated to people. While it is not the only thing that matters it seems kind of pointless to go to such lengths to hide it.
- Sometimes points are used for things that are NOT samples which can get confusing and honestly I dont like it.

# Future work/Conclusion
This paper has identified several issues in the uncertainty visualisation literature and with those issues, we have several suggestions for future work that would likely be fruitful for the field.

Our first suggestion for future work is a mathematical defintion of uncertainty. Specifically, a definition that has a place for all the concepts that currently sit under the "uncertainty" umbrella, explains how these concepts relate to uncertianty, and also allows us to quantify them in a way where they can be combined or visualised. This work would not only benefit uncertainty visualisaiton, but also statistics more broadly. There is currently a lot of work that quantifies the uncertainty through bias and variance at various stages of the data analysis pipeline separately. Elements such as imputed data, assumptions, sampling methods, and analyst decisions all have their own independent work, quantification and discussions, but methods to brand between these concepts are few and far between. A conceptual framework that allows us to combine these results would be incredibly beneficial to the field.

The concept of uncertainty should also be formalised within the grammar of graphics. @Leland2005 referred to uncertainty as "semantics", however the depth of integration required to understand problems in uncertainty visualisation suggest they *need* to be formally placed within the framework. Displaying a distribution is not as simple as displaying the data. For example, the distinction between a sample and a smothed density function is not cut and dry. One may view them both as the same data but one is represented as a point, the other is binned according to some function and then represented as a line. However, this distinction blurs as n gets larger and larger and, depending on the collision modifier used, depictions of the sample will appear identical to depictions of the mass. This is obviously true due to large sample theory, the backbone of statistics. The issue of blurred elements of the grammar of graphics in the case of distribution visualisations does not end here. Visualisations that depict mass such as a confidence interval, a boxplot, a PDF, a letterbox plot, a violin plot, a histogram, a quantile dot plot, etc, all show some version of the mass of a variable at different resolutions, but whether these veriables differ in something as low down in the grammar pipeline as the variable stage, or later at the geometry stage is unclear. The lack of formalisation in this way allows authors to compare graphics that *seem* to be similar mathematically but are actually conveying very different information. Not only would a formalisation of uncertainty visualisation within the grammar of graphics allow us to iron out some of these confusions, but they would make it easier to understand when existing visualisation methods apply and can be used to explain our results.

```{r}
# visualisation of a dot plot with increasing sample size
```

If we are going to consider uncertainty as noise, not signal, there needs to be a way to identify this signal supression in an experimental design. It is clear that value retrival and questions with a specific ground truth are not appropriate, as uncertainty visualisation will simply follow the existing visualisation rules. Testing a secondary feature of a plot, that is, how strong a signal is becomes a little more complicated. Line up protocols have been used in adjacent work that look at the strength of the signal in a plot, and this idea of identifying if plots have some "barely noticiable differences" could be utilised. There is also the possibility that uncertainty visualisation evaluations will need to swap to a qualitative methodology where participants are allowed to freely comment on what they notice in graphics until we establish how the existence of noise can be observed.

Additionally, information visualisation needs a more precise concept of the difference between two plots, and that framework should be utilised in experiments. Other fields of science employ marginal changes when designing experiments to ensure it is well understood *what* aspect of their experiment is contributing to their results. Visualisation has the grammar of graphics, however we have already discussed how this can break down in the case of distribution visualisation.

If a visualisation researcher would prefer to perform experiments rather than formalise methods, the task dependency many authors in uncertainty visualisation mention would be a useful direction for research. Unfortunately the current visualisation evaluation literature is *far* too noisy to allow us to identify what might be the cause of this task dependence. It is unclear if it is a conflation between task and the ground truth/displayed statistics concepts discussed earlier, or an actual difference in the way we percieve graphics. A good starting place for this idea would be to identify if there is task dependency in perceptual tasks, since all perceptual task research to date has been done on value retrival. That is, an experiment that compares things like area, position, colour and other perceptual tasks on value exaction, comparisons, and other simple tasks. It is also clear that the the number of potential tasks that can be performed on a visualisation increases with with the number of observations. A single observation is limited to value extraction, two observations can be compared, multiple observations allow for shapes or global statistics to be extracted. The interaction between sample size and task is of particular interest to the uncertainty visualisation community, as uncertainty can be expressed through multiple observations using a sample, or through a single value using an error. Of course, this is limited by the fact that there also isnt a definition for what is a "task" and given the mess created by the lack of formalisation in uncertainty visualisation, it may be wise to formalise that concept before performing these experiemnts. Reguardless, these are directions of research would be fruitful to the uncertainty visualisation community even if it appears on the surface to be research that is only beneficial to the "normal" visualisation community.

### 2024-02-23 writingdumpster.qmd
finished the task wordvomit and future work  sections

# Moved over notes
- mentions uncertainty can be variance, precision, accuracy, reliability, or related concepts @Hullman2016
-  @Spiegelhalter2017 defines risk as risk known probabilities and uncertainty as estimated probabilities, therefore uncertainty is what we have when you cannot accurately quantify risk.
- Authors provide little justification for their chosen response models (e.g. absolute accuracy vs relative measures) @Hullman2016 
- Some questions (e.g. what are the chances that the no.6 bus will arrive first) you can elicit the viewers subjective probability distribution @Hullman2016
- The big data paradox shrinks confidence intervals but magnifies bias [@Bradley2021] because the confidence intervals shrink due to the sample size but the imprecise target group creates bias.
- coloured maps for earthquake risk are easily interpretable @Spiegelhalter2017
- Hurricane risk is not only the path, but also the storm surge and wind speed. @Spiegelhalter2017

>>>>

# Moved over notes
- Error is best mapped to fuzziness, location, and colour value; arrangement, size and transparency are an OK second choice; but saturation, hue, orientation and shape are unacceptable and have no intuitive connection to variance [@Maceachren2012]. 
- @Nathonours found that a scatter plot is better than a line plot if you want to convey the correlation between two time series but we cannot be sure if this was influenced by swapping the distribution depicted or by dropping the irrelevant feature of inexchangeability.
- No only do the graphical elements we map our features to matter, but the direction matters too. Graphical elements that are more fuzzy (fuzziness), further from centre (location), lighter (colour value), poorly arranged (arrangement), smaller (size), more transparent (transparency) are perceived to be more uncertain [@Maceachren2012]. 
- This idea also extends to interval estimation, where questions about probability are best answered with gradients and questions about start and end times are easiest to answer with ambiguation [@Gschwandtnei2016].
- Heuristics can work against us just as much as they can work for us. The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015]
- mentions uncertainty can be variance, precision, accuracy, reliability, or related concepts @Hullman2016
-  @Spiegelhalter2017 defines risk as risk known probabilities and uncertainty as estimated probabilities, therefore uncertainty is what we have when you cannot accurately quantify risk.
- Authors provide little justification for their chosen response models (e.g. absolute accuracy vs relative measures) @Hullman2016 
- Some questions (e.g. what are the chances that the no.6 bus will arrive first) you can elicit the viewers subjective probability distribution @Hullman2016
- The big data paradox shrinks confidence intervals but magnifies bias [@Bradley2021] because the confidence intervals shrink due to the sample size but the imprecise target group creates bias.
- coloured maps for earthquake risk are easily interpretable @Spiegelhalter2017
- Hurricane risk is not only the path, but also the storm surge and wind speed. @Spiegelhalter2017

# Experiment paper checklist
General
- Specific Field of Research
- Question asked
- Risk or Uncertainty: 
- Noise vs signal (risk vs uncertainty)
- Major task goal
    - e.g. identify a bias or misunderstanding, compare two visualisation methods
- Minor task goal
    - Communication(value retrieval)
    - Analytical(complex consideration)
    - Exploration(free for all)

Uncertainty Considerations
	- Source
	- Level (statistical-ignorance)
	- Nature (Epistemic/Aleatory)

Visual Uncertainty taxonomy
- Explicit/implicit
    - (directly mapping or showing multiple outcomes)
- Intrinsic/Extrinsic 
    - (using existing symbols e.g. colour value, or new objects e.g. grids)
- Visually integral/separable 
    - (can or cannot be separated from the data) 
- Coincidence/adjacent 
    - (if data and uncertainty are in integrated or separate views
- Static/dynamic 
    - (animation/interaction)

Other plot considerations
	- Dimension of data
	- Dimension of uncertainty
	- Feature mappings

Extra (not necessarily recorded)
	- Metrics used (and recorded)
	- Possible Heuristics
	- What is the ground truth?
	- Participant literacy
